---
title: "GEMMA trans+cis-eQTLs"
author: "Daniel Munro"
date: "6/3/2021"
output: html_document
---

I'm doing all-vs-all association testing. I pruned SNPs to LD <= 0.9, resulting in 30,271 tested SNPs. I've kept only pairs with p-values < 1e-3.

<!-- For now I ran GEMMA for genes on the four chromosomes with the fewest genes (11, 15, 17, 18), for a total of 1873 genes. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
suppressPackageStartupMessages(library(tidyverse))

genes <- read_tsv("~/br/data/genes.txt", col_types = "c-c---i-----") |>
    rename(gene_chrom = chrom)

top_lm <- read_tsv("~/br/data/gemma/trans/Acbc.lm.assoc.txt.gz", col_types = "c-c--------d")

top_lmm <- read_tsv("~/br/data/gemma/trans/Acbc.lmm.assoc.txt.gz", col_types = "c-c----------d--")

top <- full_join(
    top_lm |> rename(p_lm = p_wald),
    top_lmm |> rename(p_lmm = p_wald),
    by = c("gene_id", "rs")
) |>
    separate(rs, c("chrom", "pos"), sep=":", convert = TRUE, remove = FALSE) |>
    mutate(chrom = str_replace(chrom, "chr", "")) |>
    left_join(genes, by = "gene_id") |>
    mutate(type = if_else(gene_chrom == chrom & abs(pos - tss) < 5e6, "cis", "trans"))

top2 <- top |>
    filter(!is.na(p_lm),
           !is.na(p_lmm))
```

Confirming cis-eQTL TSS distances make sense:

```{r}
top |>
    filter(type == "cis") |>
    group_by(gene_id) |>
    arrange(p_lmm) |>
    slice(1) |>
    ungroup() |>
    mutate(tss_distance = (pos - tss) / 1e6) |>
    # filter(abs(tss_distance) < 1) |>
    ggplot(aes(x = tss_distance)) +
    geom_histogram(bins = 50) +
    xlab("TSS distance (Mb)") +
    ggtitle("Top cis-eQTL per eGene (at cutoff 1e-3)")
```

Considering all SNP-gene pairs with p-value < 1e-3 in LM or LMM, counting how many have p-value < 1e-3 in one or the other or both, and stratified by cis or trans:

```{r}
with(top, table(low_p_lm = !is.na(p_lm), low_p_lmm = !is.na(p_lmm), type))
```

For both cis and trans, there is large overlap even in terms of exact SNP-gene pairs. I'll compare p-values for LM vs LMM for each pair in this overlap, i.e. look at one quadrant of the p-value scatter plot:

```{r}
top_cor <- top2 |>
    with(c(cor_p = cor(p_lm, p_lmm), cor_logp = cor(-log10(p_lm), -log10(p_lmm))))
top2 |>
    ggplot(aes(x = p_lm, y = p_lmm)) +
    geom_point(size = 0.25, alpha = 0.25) +
    coord_fixed() +
    theme_minimal() +
    ggtitle(str_glue("R = {top_cor['cor_p'] |> round(3)}"))

top2 |>
    ggplot(aes(x = -log10(p_lm), y = -log10(p_lmm))) +
    geom_point(size = 0.25, alpha = 0.25) +
    coord_fixed() +
    theme_minimal() +
    ggtitle(str_glue("R = {top_cor['cor_logp'] |> round(3)}"))
```

And stratified into cis and trans:

```{r}
tmp <- top2 |>
    group_by(type) |>
    summarise(
        n = n(),
        cor_p = cor(p_lm, p_lmm) |> round(3),
        cor_logp = cor(-log10(p_lm), -log10(p_lmm)) |> round(3)
    ) |>
    mutate(label_p = str_glue("R = {cor_p}, n = {n}"),
           label_logp = str_glue("R = {cor_logp}, n = {n}"))
top2 |>
    ggplot(aes(x = p_lm, y = p_lmm)) +
    facet_wrap(~ type) +
    geom_point(size = 0.25, alpha = 0.25) +
    geom_text(aes(label = label_p, x = 0.00025, y = 0.001), data = tmp) +
    coord_fixed() +
    theme_minimal()

top2 |>
    ggplot(aes(x = -log10(p_lm), y = -log10(p_lmm))) +
    facet_wrap(~ type) +
    geom_point(size = 0.25, alpha = 0.25) +
    geom_text(aes(label = label_logp, x = 10, y = 24), data = tmp) +
    coord_fixed() +
    theme_minimal()
```

Does LMM tend to shift p-values higher/lower?

```{r}
top2 |>
    pivot_longer(cols = c(p_lm, p_lmm), names_to = "mode", values_to = "pval", names_prefix = "p_") |>
    ggplot(aes(x = mode, y = pval)) +
    facet_wrap(~ type) +
    geom_violin() +
    geom_boxplot(width = 0.1)
top2 |>
    pivot_longer(cols = c(p_lm, p_lmm), names_to = "mode", values_to = "pval", names_prefix = "p_") |>
    ggplot(aes(x = mode, y = -log10(pval))) +
    facet_wrap(~ type) +
    geom_violin() +
    geom_boxplot(width = 0.1)
```

## QQ plots

I don't think I can correctly plot QQ against theoretical without full p-value distributions. But I can plot them against each other to compare their distributions.

```{r}
# top2 %>%
#     ggplot(aes(sample = p_lmm)) +
#     geom_qq(distribution = qunif)
# 
# top2 %>%
#     ggplot(aes(sample = -log10(p_lmm))) +
#     geom_qq(distribution = \(p) -log10(qunif(p)))
# 
# tmp <- top2 %>%
#     summarise(observed = -log10(sort(p_lm)),
#               expected = -log10(ppoints(length(p_lm)) / 1e3))
# ggplot(tmp, aes(x = expected, y = observed)) +
#     geom_point() +
#     geom_abline(slope = 1, intercept = 0) +
#     coord_fixed(xlim = c(0, max(c(tmp$observed, tmp$expected))),
#                 ylim = c(0, max(c(tmp$observed, tmp$expected)))) +
#     theme_minimal()

tmp <- top2 %>%
    group_by(type) %>%
    summarise(lm = -log10(sort(p_lm)),
              lmm = -log10(sort(p_lmm)),
              .groups = "drop")
ggplot(tmp, aes(x = lm, y = lmm)) +
    facet_wrap(~ type) +
    geom_point(size = 0.5) +
    geom_abline(slope = 1, intercept = 0, size = 0.5, color = "red", lty = 2) +
    coord_fixed(xlim = c(0, max(c(tmp$lm, tmp$lmm))),
                ylim = c(0, max(c(tmp$lm, tmp$lmm)))) +
    theme_minimal()
```

Those are for associations where LM and LMM have p<1e-3, which could be considered selecting against disparate p-values. Comparing the sets of LM and LMM p-values < 1e-3, which are of different size and thus require interpolation/subsampling for one set:

```{r}
# # Interpolate LMM values:
# tmp <- top %>%
#     group_by(type) %>%
#     summarise(lm = -log10(sort(p_lm[!is.na(p_lm)])),
#               lmm = -log10(quantile(p_lmm[!is.na(p_lmm)], ppoints(sum(!is.na(p_lm))))),
#               .groups = "drop")
# # Interpolate LM values:
# tmp <- top %>%
#     group_by(type) %>%
#     summarise(lmm = -log10(sort(p_lmm[!is.na(p_lmm)])),
#               lm = -log10(quantile(p_lm[!is.na(p_lm)], ppoints(sum(!is.na(p_lmm))))),
#               .groups = "drop")
# Subsample LMM values:
tmp <- top %>%
    mutate(type = fct_recode(type,
                             "Proximal SNP-gene pairs" = "cis",
                             "Distal SNP-gene pairs" = "trans")) |>
    group_by(type) %>%
    summarise(lm = -log10(sort(p_lm[!is.na(p_lm)])),
              lmm = -log10(sort(sample(p_lmm[!is.na(p_lmm)], sum(!is.na(p_lm))))),
              .groups = "drop")
ggplot(tmp, aes(x = lm, y = lmm)) +
    facet_wrap(~ type) +
    geom_abline(slope = 1, intercept = 0, size = 0.5, color = "red", lty = 2) +
    geom_point(size = 0.5) +
    coord_fixed(xlim = c(0, max(c(tmp$lm, tmp$lmm))),
                ylim = c(0, max(c(tmp$lm, tmp$lmm)))) +
    theme_minimal() +
    xlab(expression(-log[10]*"P, fixed effects model")) +
    ylab(expression(-log[10]*"P, linear mixed model"))

# Now plot this figure using random associations in another script.
# ggsave("gemma_fig_QQ.png", width = 5.2, height = 3, bg = "white")
```

