---
title: "GTEx tissue specificity"
output: html_notebook
---

How much do GTEx eGenes overlap among the brain tissues compared to the rat brain tissues?

```{r}
library(tidyverse)

gtex <- tibble(file = list.files("../data/gtex/GTEx_Analysis_v8_eQTL", "Brain_.+",
                                 full.names = TRUE)) %>%
    mutate(tissue = str_match(file, "Brain_(.+)\\.v8")[, 2]) %>%
    group_by(file, tissue) %>%
    summarise(
        read_tsv(file, col_types = cols(gene_id = "c", qval = "d", .default = "-")),
        .groups = "drop"
    ) %>%
    filter(qval <= 0.05) %>%  # GTEx site says to do <=
    select(-file, -qval)

rats <- read_tsv("../data/eqtls.txt", col_types = "cc----d---") %>%
    filter(qval < 0.05) %>%
    select(tissue, gene_id)

eqtls <- bind_rows(
    gtex %>% mutate(data = "GTEx"),
    rats %>% mutate(data = "rats")
)
```

```{r}
# gtex %>%
#     ggplot(aes(x = tissue)) +
#     geom_bar() +
#     coord_flip() +
#     ggtitle("GTEx brain tissues")
# 
# rats %>%
#     ggplot(aes(x = tissue)) +
#     geom_bar() +
#     coord_flip() +
#     ggtitle("HS rat brain tissues")

eqtls %>%
    mutate(data = fct_rev(data),
           tissue = fct_rev(tissue)) %>%
    ggplot(aes(x = tissue)) +
    facet_grid(rows = "data", scales = "free_y", space = "free_y") +
    geom_bar() +
    coord_flip() +
    theme_minimal() +
    xlab(NULL) +
    ylab("Number of eGenes")

# ggsave("eGene_count.png", width = 6, height = 3.5)
```

To compare eGene overlap to overlap in our data, measure pairwise overlap as Jaccard/Tanimoto index in both datasets and compare the two distributions.

```{r}
overlap <- bind_rows(
    crossing(data = "GTEx",
             tissue.x = unique(gtex$tissue),
             tissue.y = unique(gtex$tissue)),
    crossing(data = "rats",
             tissue.x = unique(rats$tissue),
             tissue.y = unique(rats$tissue))
) %>%
    filter(tissue.x < tissue.y) %>%
    group_by(data, tissue.x, tissue.y) %>%
    summarise({
        set1 <- eqtls$gene_id[eqtls$tissue == tissue.x]
        set2 <- eqtls$gene_id[eqtls$tissue == tissue.y]
        tibble(n_intersection = length(intersect(set1, set2)),
               n_union = length(union(set1, set2))) %>%
            mutate(jaccard = n_intersection / n_union)
    }, .groups = "drop")

overlap %>%
    ggplot(aes(x = data, y = jaccard)) +
    geom_boxplot(outlier.shape = NA) +
    ggbeeswarm::geom_beeswarm(size = 1, color = "blue")
```

Overlap in rat brain tissues tends to be higher, despite having overall smaller fractions of total genes, which would lead to lower overlap by chance.

As a simplistic correction for that, subset the GTEx tissues to the 5 with the fewest eGenes.

```{r}
gtex_subset <- gtex %>%
    count(tissue, sort = TRUE) %>%
    slice_tail(n = 5)

overlap %>%
    filter(data == "rats" | (tissue.x %in% gtex_subset$tissue &
                                 tissue.y %in% gtex_subset$tissue)) %>%
    ggplot(aes(x = data, y = jaccard)) +
    geom_boxplot(outlier.shape = NA) +
    ggbeeswarm::geom_beeswarm(size = 1, color = "blue")
```

```{r}
overlap %>%
    # filter(data == "rats") %>%
    ggplot(aes(x = n_union, y = n_intersection, color = data)) +
    geom_point(size = 1, alpha = 0.5) +
    expand_limits(x = 0, y = 0) +
    theme_minimal() +
    coord_fixed() +
    xlab("Size of eGene union between tissues") +
    ylab("Size of eGene intersection between tissues") +
    ggtitle("Overlap between pairs of brain tissues")

# ggsave("eGene_overlap.png", width = 6, height = 3.5)
```

## aFC

Do eGenes found in more tissues have different effect sizes than those in fewer tissues?

```{r}
gtex_afc <- tibble(file = list.files("../data/gtex/GTEx_Analysis_v8_eQTL", "Brain_.+",
                                 full.names = TRUE)) %>%
    mutate(tissue = str_match(file, "Brain_(.+)\\.v8")[, 2]) %>%
    group_by(file, tissue) %>%
    summarise(
        read_tsv(file, col_types = cols(gene_id = "c", pval_beta = "d", qval = "d",
                                        log2_aFC = "d", .default = "-")),
        .groups = "drop"
    ) %>%
    filter(qval <= 0.05) %>%  # GTEx site says to do <=
    select(-file, -qval)

rats_afc <- read_tsv("../data/eqtls.txt", col_types = "cc---dd-d-") %>%
    filter(qval < 0.05) %>%
    select(tissue, gene_id, pval_beta = pval, log2_aFC = log2_aFC_eQTL)

afc <- bind_rows(
    gtex_afc %>% mutate(data = "GTEx"),
    rats_afc %>% mutate(data = "rats")
)
```

```{r}
afc %>%
    filter(data == "rats") %>%
    group_by(gene_id) %>%
    mutate(n_tissues = n()) %>%
    ungroup() %>%
    mutate(n_tissues = as.factor(n_tissues),
           log10_pval = -log10(pval_beta)) %>%
    sample_n(nrow(.)) %>%
    ggplot(aes(x = log2_aFC, y = log10_pval, color = n_tissues)) +
    facet_wrap(~ tissue) +
    geom_point(size = 0.5, alpha = 0.75) +
    scale_color_viridis_d() +
    theme_minimal()

# ggsave("volcano_aFC.png", width = 8, height = 6)
```

```{r}
afc %>%
    filter(data == "GTEx") %>%
    group_by(gene_id) %>%
    mutate(n_tissues = n()) %>%
    ungroup() %>%
    mutate(log10_pval = -log10(pval_beta)) %>%
    sample_n(nrow(.)) %>%
    ggplot(aes(x = log2_aFC, y = log10_pval, color = n_tissues)) +
    facet_wrap(~ tissue) +
    geom_point(size = 0.5, alpha = 0.75) +
    scale_color_viridis_c() +
    theme_minimal()

# ggsave("volcano_aFC_GTEx.png", width = 8, height = 8)
```

