---
title: "gene_expr_pred"
author: "Daniel Munro"
format:
  html:
    embed-resources: true
editor: visual
execute:
  echo: false
---

I used `gene_expr_pred` to predict expression on the full NAcc dataset, which was also used to compute the aFC values (i.e. training and testing on the same data for now).

```{r}
suppressPackageStartupMessages(library(tidyverse))

tissues <- c("IL", "LHb", "NAcc", "OFC", "PL")

pred <- tibble(tissue = tissues) |>
    reframe(
        read_tsv(str_glue("../../data/expr_var_explained/gene_expr_pred/out_{tissue}/expressions/gene_expression.txt"),
                 col_types = cols(gene_id = "c", .default = "d")) |>
            pivot_longer(-gene_id, names_to = "sample", values_to = "prediction"),
        .by = tissue
    )

expr <- tibble(tissue = tissues) |>
    reframe(
        read_tsv(str_glue("../../data/expression/ensembl-gene_log2_{tissue}.bed.gz"),
                 col_types = cols(`#chr` = "-", start = "-", end = "-",
                                  gene_id = "c", .default = "d")) |>
            pivot_longer(-gene_id, names_to = "sample", values_to = "expression"),
        .by = tissue
    ) |>
    inner_join(pred, by = c("tissue", "gene_id", "sample"))

corrs <- expr |>
    reframe(
        corr = if (sd(prediction) == 0) NA else cor(expression, prediction),
        R2 = 1 - var(prediction - expression) / var(expression), # From Nava
        .by = c(tissue, gene_id)
    ) |>
    arrange(tissue, desc(R2))

write_tsv(corrs, "../../data/expr_var_explained/expr_R2.tsv")
```

```{r}
corrs |>
    ggplot(aes(x = corr)) +
    facet_wrap(~ tissue) +
    geom_histogram(bins = 30)
corrs |>
    ggplot(aes(x = R2)) +
    facet_wrap(~ tissue) +
    geom_histogram(bins = 30)
```

```{r}
corrs |>
    mutate(corr_sq = corr ^ 2, .after = corr) |>
    summary()
```

```{r fig.width = 5, fig.height = 5}
expr |>
    filter(gene_id %in% corrs$gene_id[1:9]) |>
    ggplot(aes(x = expression, y = prediction, color = tissue)) +
    facet_wrap(vars(gene_id), ncol = 3, scales = "free") +
    geom_point(size = 1, alpha = 0.5) +
    ggtitle("Best-predicted genes")
```

## Comparison to human

I got variance explained values from Nava for GTEx Adipose tissue.

```{r}
human <- tibble(n_eQTLs = 1:13) |>
    reframe({
        d <- read_csv(str_glue("../../data/expr_var_explained/Fig_1_F/Residual_info_{n_eQTLs}_variants.csv"),
                      col_types = c(gene_id = "c", .default = "d"))
        tibble(gene_id = d[["gene_id"]],
               R2_top = 1 - d[["Residual_1variant"]],
               R2_all = 1 - d[[str_glue("Residual_{n_eQTLs}variant")]])
    }, .by = n_eQTLs)
```

```{r}
human |>
    pivot_longer(c(R2_top, R2_all), names_to = "eQTLs_used", values_to = "R2") |>
    mutate(n_eQTLs = as.factor(n_eQTLs)) |>
    ggplot(aes(x = n_eQTLs, y = R2, color = eQTLs_used)) +
    geom_boxplot(outlier.size = 0.25) +
    coord_cartesian(ylim = c(0, 1))
```

```{r}
tmp <- bind_rows(
    corrs |>
        select(dataset = tissue, R2),
    human |>
        select(R2 = R2_top) |>
        mutate(dataset = "GTEx Adipose, top eQTL",
               R2 = pmax(-0.5, R2)),
    human |>
        select(R2 = R2_all) |>
        mutate(dataset = "GTEx Adipose, all eQTLs")
)
tmp |>
    ggplot(aes(x = dataset, y = R2)) +
    geom_violin()
```

Boxen plots (Widest box is inter-quartile range, and each additional box shows half of remaining values):

```{r}
tmp |>
    ggplot(aes(x = dataset, y = R2)) +
    lvplot::geom_lv(fill = "white", color = "black")
```

```{r}
tmp |>
    group_by(dataset) |>
    skimr::skim()
```

## Subset to homolog pairs

Comparing rat R2 to human R2 using all eQTLs, since I'm using all eQTLs for rat:

```{r}
# Made in rat_human_aFC_stats.R:
gene_map <- read_tsv("../../data/afc/gene_map.txt", col_types = "cc") |>
    mutate(gene_id_human = str_replace(gene_id_human, "\\..+$", ""))

hom_r2 <- tibble(tis = tissues) |>
    reframe(
        gene_map |>
            rename(gene_id.rat = gene_id_rat,
                   gene_id.human = gene_id_human) |>
            inner_join(
                corrs |>
                    filter(tissue == tis) |>
                    select(gene_id.rat = gene_id,
                           R2.rat = R2),
                by = "gene_id.rat"
            ) |>
            inner_join(
                human |>
                    select(gene_id.human = gene_id,
                           R2.human = R2_all),
                by = "gene_id.human"
            ),
        .by = tis
    )

tmp2 <- hom_r2 |>
    select(tis, R2.rat, R2.human) |>
    pivot_longer(c(R2.rat, R2.human), names_to = "organism", names_prefix = "R2.",
                 values_to = "R2")

tmp2 |>
    ggplot(aes(x = organism, y = R2)) +
    facet_wrap(~ tis, nrow = 1) +
    geom_violin() +
    geom_boxplot(width = 0.1, outlier.shape = NA)
```

```{r}
tmp2 |>
    group_by(tis, organism) |>
    skimr::skim()
```
