---
title: "aFC across tissues"
author: "Daniel Munro"
format:
  html:
    embed-resources: true
editor: visual
execute:
  echo: false
---

```{r}
suppressPackageStartupMessages(library(tidyverse))

tissues <- c("IL", "LHb", "NAcc", "OFC", "PL")

expr <- tibble(tissue = tissues) |>
    reframe(
        read_tsv(str_glue("../../data/expression/ensembl-gene_log2_{tissue}.bed.gz"),
                 col_types = cols(`#chr` = "-", start = "-", end = "-",
                                  gene_id = "c", .default = "d")) |>
            pivot_longer(-gene_id, names_to = "rat_id", values_to = "expr"),
        .by = tissue
    ) |>
    summarise(median_expr = median(expr),
              frac_nonzero = mean(expr > 0),
              .by = c(tissue, gene_id))

# Genes with consistent expression
consist_expr <- expr |>
    group_by(gene_id) |>
    filter(min(median_expr) >= 0.5 * max(median_expr)) |>
    ungroup() |>
    distinct(gene_id) |>
    pull()

eqtls_all <- read_tsv("../../data/eqtls/top_assoc.txt", col_types = "ccciicciccdddddddd")

eqtls <- filter(eqtls_all, qval < 0.05)

afc <- tibble(tissue = tissues) |>
    reframe(
        read_tsv(str_glue("../../data/specificity/{tissue}.aFC.txt"),
                 col_types = "cc--d--") |>
            rename(gene_id = pid,
                   variant_id = sid),
        .by = tissue
    ) |>
    filter(gene_id %in% consist_expr,
           gene_id %in% eqtls$gene_id) |>
    # Note which gene/SNP pairs are eQTLs in each tissue:
    left_join(eqtls |>
                  select(tissue, gene_id, variant_id) |>
                  mutate(eQTL = TRUE),
              by = c("tissue", "gene_id", "variant_id")) |>
    replace_na(list(eQTL = FALSE))
```

Compare aFC consistency to eQTL tissue specificity per gene, doing pairwise comparisons between tissues. For each tissue pair, for each gene, get absolute difference in aFC (if tissues have different top SNP, average the two). For each tissue pair, compare distributions for genes in which both tissues vs. only one tissue has eQTLs.

```{r}
pairs <- crossing(tissue1 = tissues,
                  tissue2 = tissues,
         gene_id = unique(afc$gene_id)) |>
    left_join(afc, by = c("tissue1" = "tissue", "gene_id"),
              relationship = "many-to-many") |>
    left_join(afc, by = c("tissue2" = "tissue", "gene_id", "variant_id"),
              relationship = "many-to-one") |>
    filter(eQTL.x | eQTL.y) |>
    mutate(diff = abs(log2_aFC.x - log2_aFC.y),
           avg = (log2_aFC.x + log2_aFC.y) / 2) |>
    summarise(mean_diff = mean(diff),
              mean_avg = mean(avg),
              shared = any(eQTL.x) & any(eQTL.y),
              .by = c(tissue1, tissue2, gene_id)) |>
    filter(is.finite(mean_diff))
```

```{r}
pairs |>
    ggplot(aes(y = mean_diff, x = shared)) +
    facet_grid(rows = vars(tissue1), cols = vars(tissue2)) +
    geom_boxplot(outlier.size = 0.2) +
    coord_cartesian(ylim = c(0, 1)) +
    xlab("eQTL in both tissues rather than only one") +
    ylab("Absolute difference in aFC")
```

eQTLs found in both tissues seem to have slightly but consistently lower effect size difference than those found in only one of the two tissues. Combining all unique tissue pairs:

```{r}
pairs |>
    filter(tissue1 < tissue2) |>
    ggplot(aes(y = mean_diff, x = shared)) +
    geom_violin(color = "gray", alpha = 0.5) +
    geom_boxplot(outlier.size = 0.25, width = 0.4) +
    scale_y_log10() +
    xlab("eQTL in both tissues rather than only one") +
    ylab("Absolute difference in aFC")
```

```{r}
pairs |>
    mutate(abs_mean_avg = abs(mean_avg)) |>
    filter(tissue1 < tissue2) |>
    ggplot(aes(y = abs_mean_avg, x = shared)) +
    geom_violin(color = "gray", alpha = 0.5) +
    geom_boxplot(outlier.size = 0.25, width = 0.4) +
    xlab("eQTL in both tissues rather than only one") +
    ylab("Absolute average aFC")
```

```{r}
eqtls |>
    mutate(log_pval = -log10(pval_beta)) |>
    ggplot(aes(x = log2_aFC, y = log_pval)) +
    geom_point(size = 0.25, alpha = 0.5) +
    stat_smooth()
```

```{r}
tmp <- eqtls_all |>
    mutate(log_pval = -log10(pval_beta),
           eQTL = qval < 0.05) |>
    group_by(gene_id) |>
    filter(n() == 5) |>
    mutate(n_tissues = sum(eQTL),
           med_logp = median(log_pval)) |>
    ungroup() |>
    filter(n_tissues > 0)
tmp |>
    mutate(n_tissues = as_factor(n_tissues)) |>
    ggplot(aes(x = n_tissues, y = log_pval, color = eQTL)) +
    geom_jitter(size = 0.25, alpha = 0.5) +
    geom_boxplot(alpha = 0.5, outlier.shape = NA)
```

```{r}
tmp |>
    mutate(gene_rank = gene_id |> fct_reorder(med_logp) |> as.integer()) |>
    ggplot(aes(x = gene_rank, y = log_pval, color = eQTL)) +
    geom_point(size = 0.25, alpha = 0.5) +
    xlab("eGenes ranked by median -log10(p-value) of all 5 tissues")
```

```{r}
# Save gene lists for eGenes in 1 tissues, 5 tissues, and all genes to do GO enrichment.
eqtls_all |>
    distinct(gene_id) |>
    pull() |>
    write_lines("../../data/specificity/GO/all_genes.txt")

tmp |>
    filter(n_tissues == 1) |>
    distinct(gene_id) |>
    pull() |>
    write_lines("../../data/specificity/GO/single_tissue.txt")

tmp |>
    filter(n_tissues == 5) |>
    distinct(gene_id) |>
    pull() |>
    write_lines("../../data/specificity/GO/all_tissues.txt")
```

There are actually very few zero expression counts:

```{r}
ggplot(expr, aes(x = frac_nonzero)) +
    geom_histogram(bins = 100)
```

```{r}
expr |>
    filter(gene_id %in% eqtls$gene_id) |>
    summarise(min_frac_nonzero = min(frac_nonzero),
              .by = gene_id) |>
    summarise(frac = mean(min_frac_nonzero < 0.5))
```

```{r}
expr |>
    filter(gene_id %in% eqtls$gene_id) |>
    summarise(foldchange = max(median_expr) / min(median_expr),
              .by = gene_id) |>
    summarise(frac_inconsistent = mean(foldchange > 2, na.rm = TRUE))
```

