---
title: "eGene pLI"
output: html_notebook
---

```{r}
library(tidyverse)

pli <- read_tsv("data/human/forweb_cleaned_exac_r03_march16_z_data_pLI.txt.gz",
                col_types = "-c-----------------d")

rat <- read_tsv("data/eqtls/top_assoc.txt", col_types = "cc-------------d--") |>
    group_by(gene_id) |>
    summarise(eqtl_rat = any(qval < 0.05),
              tissues_rat = sum(qval < 0.05),
              .groups = "drop")

gtex <- list.files("data/gtex/GTEx_Analysis_v8_eQTL", pattern = "*Brain_*",
                   full.names = TRUE) |>
    read_tsv(id = "file",
             col_types = cols(gene_id = "c", gene_name = "c", qval = "d", .default = "-")) |>
    mutate(gene_id = str_replace(gene_id, "\\..+$", "")) |>
    group_by(gene_id, gene_name) |>
    summarise(eqtl_gtex = any(qval <= 0.05),
              tissues_gtex = sum(qval <= 0.05),
              .groups = "drop")

d <- read_tsv("data/gtex/orthologs.txt", col_types = "cc") |>
    mutate(gene_id_human = str_replace(gene_id_human, "\\..+$", "")) |>
    inner_join(rat, by = c("gene_id_rat" = "gene_id")) |>
    inner_join(gtex, by = c("gene_id_human" = "gene_id")) |>
    inner_join(pli, by = c("gene_name" = "gene")) |>
    mutate(
        eqtl = case_when(
            eqtl_rat & !eqtl_gtex ~ "Rat",
            !eqtl_rat & eqtl_gtex ~ "GTEx",
            eqtl_rat & eqtl_gtex ~ "Both",
            !eqtl_rat & !eqtl_gtex ~ "Neither",
        ) |>
            fct_relevel("Both", "GTEx", "Rat")
    )

# expr <- tibble(tissue = c("IL", "LHb", "NAcc", "OFC", "PL")) |>
#     group_by(tissue) |>
#     summarise(
#         read_tsv(str_glue("data/expression/ensembl-gene_log2_{tissue}.bed.gz"),
#                  col_types = cols(`#chr` = "-", start = "-", end = "-",
#                                   gene_id = "c", .default = "d")) |>
#             pivot_longer(-gene_id, names_to = "rat_id", values_to = "expr"),
#         .groups = "drop"
#     ) |>
#     group_by(gene_id) |>
#     summarise(median_expr = median(expr))
expr <- read_tsv("data/expression/medianGeneExpression.txt.gz", col_types = "cddddddddd") |>
    pivot_longer(-geneId, names_to = "tissue", values_to = "tpm") |>
    filter(tissue %in% c("IL", "LHb", "NAcc", "OFC", "PL")) |>
    rename(gene_id = geneId) |>
    group_by(gene_id) |>
    summarise(n_expr = sum(tpm > 1), .groups = "drop") |>
    filter(n_expr == 5)
```

```{r}
d |>
    ggplot(aes(x = eqtl, y = pLI)) +
    geom_violin(fill = "grey") +
    geom_boxplot(width = 0.05) +
    xlab("Gene has an eQTL in any tissue from...")

with(d, table(eqtl))
```

```{r}
d |>
    group_by(tissues_rat, tissues_gtex) |>
    summarise(mean_pLI = mean(pLI),
              n = n(),
              .groups = "drop") |>
    ggplot(aes(x = tissues_rat, y = tissues_gtex, fill = mean_pLI, label = n)) +
    geom_tile() +
    geom_text(color = "white") +
    scale_fill_viridis_c() +
    ggtitle("Genes (ortho pairs) grouped by # of tissues w/ eQTL",
            subtitle = "Tiles are labeled with gene count")
```

pLI isn't highest for 0-eQTL genes. Is it because the mean is dragged down by genes with regulatory variation but have low expression so eQTLs aren't detected?

Maybe so, though note that the mean for 0 rat eQTLs is already the highest overall:

```{r}
# d |>
#     left_join(expr, by = c("gene_id_rat" = "gene_id")) |>
#     ggplot(aes(x = mean_expr, y = pLI)) +
#     facet_wrap(~ tissues_rat) +
#     geom_point(size = 1)

d |>
    mutate(tissues_rat = as.factor(tissues_rat)) |>
    # left_join(expr, by = c("gene_id_rat" = "gene_id")) |>
    ggplot(aes(x = tissues_rat, fill = gene_id_rat %in% expr$gene_id, y = pLI)) +
    # facet_wrap(~ tissues_rat) +
    geom_boxplot()
```

Removing genes with median rat TPM < 1 in all tissues:

```{r}
d |>
    # left_join(expr, by = c("gene_id_rat" = "gene_id")) |>
    filter(gene_id_rat %in% expr$gene_id) |>
    group_by(tissues_rat, tissues_gtex) |>
    summarise(mean_pLI = mean(pLI),
              n = n(),
              .groups = "drop") |>
    ggplot(aes(x = tissues_rat, y = tissues_gtex, fill = mean_pLI, label = n)) +
    geom_tile() +
    geom_text(color = "white") +
    scale_fill_viridis_c() +
    ggtitle("Genes (ortho pairs) grouped by # of tissues w/ eQTL",
            subtitle = "Tiles are labeled with gene count")
```

Mean +/- SE instead of boxplots, filtered out low-expressed genes, and include genes that were tested in only one species:

```{r}
d2 <- read_tsv("data/gtex/orthologs.txt", col_types = "cc") |>
    mutate(gene_id_human = str_replace(gene_id_human, "\\..+$", "")) |>
    left_join(rat, by = c("gene_id_rat" = "gene_id")) |>
    left_join(gtex, by = c("gene_id_human" = "gene_id")) |>
    inner_join(pli, by = c("gene_name" = "gene")) |>
    mutate(
        eqtl = case_when(
            eqtl_rat & !eqtl_gtex ~ "Rat",
            !eqtl_rat & eqtl_gtex ~ "GTEx",
            eqtl_rat & eqtl_gtex ~ "Both",
            !eqtl_rat & !eqtl_gtex ~ "Neither",
        ) |>
            fct_relevel("Both", "GTEx", "Rat")
    )

d |>
    # left_join(expr, by = c("gene_id_rat" = "gene_id")) |>
    filter(gene_id_rat %in% expr$gene_id) |>
    group_by(eqtl) |>
    summarise(mean_pLI = mean(pLI),
              SE_pLI = sd(pLI) / sqrt(length(pLI)),
              n = n(),
              .groups = "drop") |>
    ggplot(aes(x = eqtl, y = mean_pLI, ymin = mean_pLI - SE_pLI, ymax = mean_pLI + SE_pLI)) +
    geom_pointrange() +
    xlab("Gene has an eQTL in any tissue from...")
```

