---
title: "cis-heritability"
author: "Daniel Munro"
format:
  html:
    embed-resources: true
editor: visual
execute:
  echo: false
---

```{r}
suppressPackageStartupMessages(library(tidyverse))

vg <- tibble(tissue = c("IL", "LHb", "NAcc", "OFC", "PL")) |>
    reframe(
        read_tsv(str_glue("../../data/anevah/output/Vg.{tissue}.tsv.gz"), col_types = "cd"),
        .by = tissue
    ) |>
    rename(gene_id = GENE_ID)

eqtls <- read_tsv("../../data/eqtls/eqtls_indep.txt", col_types = "ccciiciiccdddddid")
has_eqtl <- eqtls |>
    select(tissue, gene_id) |>
    mutate(eqtl = TRUE)

pve <- tibble(tissue = c("IL", "LHb", "NAcc", "OFC", "PL")) |>
    reframe(
        read_tsv(str_glue("../../data/gemma/{tissue}.cis_pve.txt"), col_types = "cd"),
        .by = tissue
    ) |>
    filter(!is.na(pve)) |>
    mutate(pve_raw = pve,
           pve = pmax(0, pmin(pve_raw, 1))) |>
    left_join(has_eqtl, by = c("tissue", "gene_id")) |>
    replace_na(list(eqtl = FALSE)) |>
    left_join(vg, by = c("tissue", "gene_id"))
```

Reported PVE:

```{r}
pve |>
    ggplot(aes(x = pve_raw)) +
    facet_wrap(~ tissue) +
    geom_histogram(bins = 100)
```

PVE clipped to [0, 1] as I can see in [A characterization of cis- and trans-heritability of RNA-Seq-based gene expression](https://www.nature.com/articles/s41431-019-0511-5):

```{r}
pve |>
    ggplot(aes(x = pve)) +
    facet_wrap(~ tissue) +
    geom_histogram(bins = 100)
```

Does cis-PVE for each gene correlate across tissues?

```{r}
pve |>
    filter(n() == 5,
           runif(1) < 0.1,
           .by = gene_id) |>
    reframe({
        df <- tibble(tissue, pve)
        crossing(tissue.x = tissue,
                 tissue.y = tissue) |>
            left_join(df, by = c("tissue.x" = "tissue"), relationship = "many-to-many") |>
            left_join(df, by = c("tissue.y" = "tissue"), relationship = "many-to-many")
    }, .by = gene_id) |>
    ggplot(aes(x = pve.x, y = pve.y)) +
    facet_grid(rows = vars(tissue.x), cols = vars(tissue.y)) +
    geom_point(size = 0.5, alpha = 0.5)
```

Standard errors are quite large:

```{r}
pve |>
    ggplot(aes(x = pve, y = se_pve, color = tissue)) +
    geom_point(size = 0.5, alpha = 0.5) +
    geom_abline(slope = 1, intercept = 0) +
    coord_fixed() +
    ggtitle("Line is x = y")
```

## Compare to eQTLs and Vg

```{r}
pve |>
    ggplot(aes(x = eqtl, y = pve)) +
    facet_grid(cols = vars(tissue)) +
    geom_boxplot() +
    geom_violin()
```

```{r}
mean_pve <- pve |>
    group_by(gene_id) |>
    filter(any(!is.na(Vg))) |>
    summarise(mean_pve = mean(pve),
              mean_Vg = mean(Vg, na.rm = TRUE))
corr <- with(mean_pve, cor(mean_pve, mean_Vg, use = "complete.obs"))
mean_pve |>
    ggplot(aes(x = mean_pve, y = mean_Vg)) +
    geom_point(size = 0.5, alpha = 0.5) +
    ggtitle(str_glue("Pearson R = {corr}"))
```

## Compare to human (NOTE: non-brain human tissues only)

The GTEx heritability estimates were produced in https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1006423 and obtained from https://github.com/jlbren/GenArchDB. They did Bayesian Sparse LMM using GCTA (h2) and LMM using GEMMA (pve).

```{r}
gene_map <- read_tsv("../../data/gtex/orthologs.txt", col_types = "cc") |>
    mutate(gene_id_human = str_replace(gene_id_human, "\\..+$", ""))

rat <- pve |>
    group_by(gene_id) |>
    summarise(mean_h2_rat = mean(pve))

gtex <- read_csv("../../data/gtex/heritability/old/genarch.csv.gz", col_types = "-ccd-d----") |>
    mutate(ensid = str_replace(ensid, "\\..+$", "")) |>
    group_by(ensid) |>
    summarise(mean_h2_human = mean(h2),
              mean_pve_human = mean(pve))

mean_ortho <- gene_map |>
    inner_join(rat, by = c("gene_id_rat" = "gene_id"), relationship = "many-to-one") |>
    inner_join(gtex, by = c("gene_id_human" = "ensid"), relationship = "many-to-one")
```

```{r}
deming <- mean_ortho |>
    slice_sample(n = 2000) |>
    summarise({
        coef <- deming::deming(mean_h2_human ~ mean_h2_rat)$coefficients
        tibble(intercept = coef[1],
               slope = coef[2])
    })

mean_ortho |>
    ggplot(aes(x = mean_h2_rat, y = mean_h2_human)) +
    geom_point(size = 0.5, alpha = 0.5) +
    geom_abline(aes(intercept = intercept, slope = slope), data = deming, color = "#8888cc") +
    coord_fixed() +
    expand_limits(x = 0, y = 0) +
    theme_bw() +
    theme(panel.grid = element_blank())
```

```{r}
deming2 <- mean_ortho |>
    slice_sample(n = 2000) |>
    summarise({
        coef <- deming::deming(mean_pve_human ~ mean_h2_rat)$coefficients
        tibble(intercept = coef[1],
               slope = coef[2])
    })

mean_ortho |>
    ggplot(aes(x = mean_h2_rat, y = mean_pve_human)) +
    geom_point(size = 0.5, alpha = 0.5) +
    geom_abline(aes(intercept = intercept, slope = slope), data = deming2, color = "#8888cc") +
    coord_fixed() +
    expand_limits(x = 0, y = 0) +
    theme_bw() +
    theme(panel.grid = element_blank())
```

```{r}
with(mean_ortho, cor.test(mean_h2_rat, mean_h2_human))
with(mean_ortho, cor.test(mean_h2_rat, mean_pve_human))
```

