---
title: "Vg from ANEVA-h"
output: 
  html_notebook: 
    code_folding: hide
---

$V^G$ is ["the expected variance in the dosage that is due to interindividual genetic differences within a population"](https://www.science.org/doi/10.1126/science.aay0256). I obtained precomputed $V^G$ estimates for GTEx (v8) and ran ANEVA-h on my phASER ASE counts to get $V^G$ per gene for HS rat brain tissues.

```{r}
suppressPackageStartupMessages(library(tidyverse))

heatmap_order <- function(fac1, fac2, value) {
    df <- tibble(fac1, fac2, value) |>
        pivot_wider(fac1, names_from = fac2, values_from = value)
    mat <- df |> select(-fac1) |> as.matrix()
    rownames(mat) <- df$fac1
    clust <- hclust(dist(mat))
    factor(fac1, levels = clust$labels[clust$order])
}

# load("../data/anevah/Vg_GTEx_v7.rda")
# gtex <- Vg_GTEx_v7 |>
# gtex <- read_tsv("../data/anevah/Vg_GTEx_v8.txt",
#                  col_types = cols(IDs = "c", .default = "d")) |>
#     select(-`...1`) |>
#     rename(gene_id = IDs) |>
#     pivot_longer(-gene_id, names_to = "tissue", values_to = "Vg")
# # rm(Vg_GTEx_v7)
gtex <- read_tsv("Vg_GTEx_phASER.tsv", col_types = cols(GENE_ID = "c", .default = "d")) |>
    rename(gene_id = GENE_ID) |>
    pivot_longer(-gene_id, names_to = "tissue", values_to = "Vg")

rat <- tibble(tissue = c("IL", "LHb", "NAcc", "OFC", "PL")) |>
    group_by(tissue) |>
    summarise(
        read_tsv(str_glue("Vg.{tissue}.tsv.gz"), col_types = "cd"),
        .groups = "drop"
    ) |>
    rename(gene_id = GENE_ID)

tissues <- jsonlite::fromJSON("../data/gtex/tissueInfo.json")[[1]] |>
    select(tissueSiteDetailAbbr,
           tissueSiteDetail) |>
    deframe() |>
    c(IL = "Infralimbic cortex",
      LHb = "Lateral habenula",
      NAcc = "Nucleus accumbens core",
      OFC = "Orbitofrontal cortex",
      PL = "Prelimbic cortex")

gene_map <- read_tsv("../data/gtex/orthologs.txt", col_types = "cc") |>
    mutate(gene_id_human = str_replace(gene_id_human, "\\..+$", ""))

vg_all <- bind_rows(
    gtex |>
        filter(!is.na(Vg)) |>
        mutate(data = if_else(str_sub(tissue, 1, 3) == "BRN", "GTEx_brain", "GTEx_other"),
               .before = "tissue"),
    rat |>
        mutate(data = "rat", .before = "tissue")
)

vg <- vg_all |>
    filter(data == "rat" | str_sub(tissue, 1, 3) == "BRN")
```

```{r}
# all_ase <- tibble(tissue = unique(rat$tissue)) |>
#     group_by(tissue) |>
#     summarise(
#         tibble(gene_id = str_glue("../data/anevah/anevah_{tissue}_ref_counts.tsv") |>
#                    read.delim(sep = "\t", row.names = 1) |>
#                    rownames()),
#         .groups = "drop"
#     ) |>
#     left_join(df, by = c("tissue", "gene_id"))
# 
# mean(!is.na(gtex$Vg))
# mean(!is.na(all_ase$Vg))
```

Number of rat genes with Vg estimates is on the low range of GTEx:

```{r fig.width = 4.5}
vg_all |>
    mutate(tissue = tissues[tissue] |> fct_infreq()) |>
    ggplot(aes(x = tissue, fill = data)) +
    geom_bar() +
    theme(axis.text.x = element_text(hjust = 1, vjust = 0.5, angle = 90))
```

Correlations in Vg for GTEx tissue pairs. Rows/cols are ordered as if there were dendrograms:

```{r fig.width = 5, fig.height = 4}
# vg |>
#     filter(data == "GTEx") |>
#     pivot_wider(gene_id, names_from = "tissue", values_from = "Vg") |>
#     select(-gene_id)
cor_gtex <- vg_all |>
    filter(data != "rat") |>
    with(crossing(tissue.x = unique(tissue),
                  tissue.y = unique(tissue),
                  gene_id = unique(gene_id))) |>
    left_join(select(vg_all, tissue.x = tissue, gene_id, Vg.x = Vg),
              by = c("tissue.x", "gene_id")) |>
    left_join(select(vg_all, tissue.y = tissue, gene_id, Vg.y = Vg),
              by = c("tissue.y", "gene_id")) |>
    filter(!is.na(Vg.x),
           !is.na(Vg.y)) |>
    group_by(tissue.x, tissue.y) |>
    summarise(r = cor(Vg.x, Vg.y),
              n = n(),
              .groups = "drop")

cor_gtex |>
    mutate(tissue.x = tissues[tissue.x],
           tissue.y = tissues[tissue.y],
           tissue.x = heatmap_order(tissue.x, tissue.y, r),
           tissue.y = heatmap_order(tissue.y, tissue.x, r)) |>
    ggplot(aes(x = tissue.x, y = tissue.y, fill = r)) +
    geom_tile() +
    coord_fixed() +
    scale_fill_viridis_c(limits = c(0, 1)) +
    xlab(NULL) +
    ylab(NULL) +
    theme(axis.text.x = element_text(hjust = 1, vjust = 0.5, angle = 90))
```

HS rat brain tissues:

```{r}
cor_rat <- vg |>
    filter(data == "rat") |>
    with(crossing(tissue.x = unique(tissue),
                  tissue.y = unique(tissue),
                  gene_id = unique(gene_id))) |>
    left_join(select(vg, tissue.x = tissue, gene_id, Vg.x = Vg),
              by = c("tissue.x", "gene_id")) |>
    left_join(select(vg, tissue.y = tissue, gene_id, Vg.y = Vg),
              by = c("tissue.y", "gene_id")) |>
    filter(!is.na(Vg.x),
           !is.na(Vg.y)) |>
    group_by(tissue.x, tissue.y) |>
    summarise(r = cor(Vg.x, Vg.y),
              n = n(),
              .groups = "drop")

cor_rat |>
    mutate(tissue.x = tissues[tissue.x],
           tissue.y = tissues[tissue.y],
           tissue.x = heatmap_order(tissue.x, tissue.y, r),
           tissue.y = heatmap_order(tissue.y, tissue.x, r)) |>
    ggplot(aes(x = tissue.x, y = tissue.y, fill = r)) +
    geom_tile() +
    coord_fixed() +
    scale_fill_viridis_c(limits = c(0, 1)) +
    # scale_fill_gradient2(limits = c(0, 1)) +
    xlab(NULL) +
    ylab(NULL) +
    theme(axis.text.x = element_text(hjust = 1, vjust = 0.5, angle = 90))
```

Compare all brain tissues from both species. For cross-species pairs compare orthologs, and for within-species pairs subset to genes that have an ortholog with Vg in at least one tissue. (The tissues that start with "Brain - " are GTEx, others are rat):

```{r}
has_vg <- gene_map |>
    filter(gene_id_rat %in% vg$gene_id,
           gene_id_human %in% vg$gene_id) |>
    with(unique(c(gene_id_rat, gene_id_human)))

cor_all <- vg |>
    with(crossing(tissue.x = unique(tissue),
                  tissue.y = unique(tissue))) |>
    group_by(tissue.x, tissue.y) |>
    summarise({
        data.x <- vg |>
            filter(tissue == tissue.x) |>
            with(unique(data))
        data.y <- vg |>
            filter(tissue == tissue.y) |>
            with(unique(data))
        if (data.x == data.y) {
            inner_join(
                filter(vg, tissue == tissue.x),
                filter(vg, tissue == tissue.y),
                by = "gene_id"
            ) |>
                filter(gene_id %in% has_vg) |>
                rename(gene_id.x = gene_id) |>
                mutate(gene_id.y = gene_id.x)
        } else if (data.x == "rat") {
            gene_map |>
                rename(gene_id.x = gene_id_rat,
                       gene_id.y = gene_id_human) |>
                inner_join(
                    filter(vg, tissue == tissue.x),
                    by = c("gene_id.x" = "gene_id")
                ) |>
                inner_join(
                    filter(vg, tissue == tissue.y),
                    by = c("gene_id.y" = "gene_id")
                )
        } else {
            gene_map |>
                rename(gene_id.x = gene_id_human,
                       gene_id.y = gene_id_rat) |>
                inner_join(
                    filter(vg, tissue == tissue.x),
                    by = c("gene_id.x" = "gene_id")
                ) |>
                inner_join(
                    filter(vg, tissue == tissue.y),
                    by = c("gene_id.y" = "gene_id")
                )
        }
    }, .groups = "drop") |>
    group_by(tissue.x, tissue.y) |>
    summarise(r = cor(Vg.x, Vg.y),
              n = n(),
              .groups = "drop")
```

```{r  fig.width = 3, fig.height = 3}
cor_all |>
    mutate(tissue.x = tissues[tissue.x],
           tissue.y = tissues[tissue.y],
           tissue.x = heatmap_order(tissue.x, tissue.y, r),
           tissue.y = heatmap_order(tissue.y, tissue.x, r)) |>
    ggplot(aes(x = tissue.x, y = tissue.y, fill = r)) +
    geom_tile() +
    coord_fixed() +
    expand_limits(fill = 0) +
    scale_fill_viridis_c() +
    # scale_fill_gradient2() +
    xlab(NULL) +
    ylab(NULL) +
    theme(axis.text.x = element_text(hjust = 1, vjust = 0.5, angle = 90))
```

Number of ortholog pairs with Vg, used to calculate the above:

```{r  fig.width = 3, fig.height = 3}
cor_all |>
    mutate(tissue.x = tissues[tissue.x],
           tissue.y = tissues[tissue.y],
           tissue.x = heatmap_order(tissue.x, tissue.y, r),
           tissue.y = heatmap_order(tissue.y, tissue.x, r)) |>
    ggplot(aes(x = tissue.x, y = tissue.y, fill = n)) +
    geom_tile() +
    coord_fixed() +
    # scale_fill_viridis_c() +
    # scale_fill_gradient() +
    xlab(NULL) +
    ylab(NULL) +
    theme(axis.text.x = element_text(hjust = 1, vjust = 0.5, angle = 90))
```

Compare average $SD^G$ ($\sqrt{V^G}$) per gene in human vs. average in rat, with Deming regression line.

```{r}
mean_ortho <- gene_map |>
    inner_join(
        vg |>
            filter(data == "rat") |>
            group_by(gene_id) |>
            summarise(SDg_rat = mean(sqrt(Vg)), .groups = "drop"),
        by = c("gene_id_rat" = "gene_id")
    ) |>
    inner_join(
        vg |>
            filter(data == "GTEx_brain") |>
            group_by(gene_id) |>
            summarise(SDg_GTEx = mean(sqrt(Vg)), .groups = "drop"),
        by = c("gene_id_human" = "gene_id")
    )

deming <- mean_ortho |>
    summarise({
        coef <- deming::deming(SDg_GTEx ~ SDg_rat)$coefficients
        tibble(intercept = coef[1],
               slope = coef[2])
    })

mean_ortho |>
    mutate(SDg_rat = pmin(SDg_rat, 0.5),
           SDg_GTEx = pmin(SDg_GTEx, 0.5)) |>
    ggplot(aes(x = SDg_rat, y = SDg_GTEx)) +
    geom_point(size = 1, alpha = 0.5) +
    geom_abline(aes(intercept = intercept, slope = slope), data = deming,
                color = "#8888cc") +
    coord_fixed() +
    ggtitle(str_glue("Pearson's r = {with(mean_ortho, cor(SDg_GTEx, SDg_rat) |> round(4))}"))
```

