---
title: "Rat-human aFC comparison"
output: html_notebook
---

```{r}
library(tidyverse)

eqtls <- read_tsv("../../data/eqtls/eqtls_indep.txt", col_types = "ccciiciiccdddddid") |>
    filter(rank == 1)

genes <- unique(eqtls$gene_id)

# # Get homologs
# # (https://www.r-bloggers.com/2016/10/converting-mouse-to-human-gene-names-with-biomart-package/)
# human <- biomaRt::useMart("ensembl", dataset = "hsapiens_gene_ensembl")
# rat <- biomaRt::useMart("ensembl", dataset = "rnorvegicus_gene_ensembl")
# gene_map <- biomaRt::getLDS(
#     attributes = c("ensembl_gene_id"),
#     filters = "ensembl_gene_id",
#     values = genes,
#     mart = rat,
#     attributesL = c("ensembl_gene_id_version"),
#     martL = human,
#     uniqueRows = TRUE
# ) |>
#     rename(gene_id = Gene.stable.ID,
#            human_id = Gene.stable.ID.version)
# write_tsv(gene_map, "../../data/afc/gene_map.txt")
gene_map <- read_tsv("../../data/afc/gene_map.txt", col_types = "cc")

gtex <- tibble(file = list.files("../../data/gtex/GTEx_Analysis_v8_eQTL",
                                 pattern = "Brain_.+",
                                 full.names = TRUE)) |>
    mutate(tissue = str_match(file, "Brain_(.+)\\.v8")[, 2]) |>
    reframe(
        read_tsv(file, col_types = cols(gene_id = "c", qval = "d", pval_beta = "-",
                                        log2_aFC = "d", .default = "-")),
        .by = c(file, tissue)
    ) |>
    filter(qval <= 0.05) |>  # GTEx site says to do <=
    select(-file)

both <- bind_rows(
    eqtls |> mutate(tissue = str_c("HSrat_", tissue)),
    gtex |> mutate(tissue = str_c("GTEx_", tissue))
)
```

```{r}
both |>
    ggplot(aes(x = tissue, y = log2_aFC)) +
    geom_boxplot(outlier.size = 0.25) +
    coord_flip()
```

The medians are interesting:

```{r}
both |>
    summarise(mean_log2_aFC = mean(log2_aFC),
              median_log2_aFC = median(log2_aFC),
              .by = tissue) |>
    pivot_longer(c(mean_log2_aFC, median_log2_aFC),
                 names_to = "stat", values_to = "value") |>
    ggplot(aes(x = tissue, y = value, color = stat)) +
    geom_hline(yintercept = 0, lty = 2) +
    geom_point() +
    coord_flip() +
    theme(legend.position = "top")
```

```{r}
gene_afc <- both |>
    group_by(gene_id) |>
    summarise(mean_abs_log2_aFC = mean(abs(log2_aFC)))
d <- gene_map |>
    inner_join(gene_afc |> rename(mean_abs_log2_aFC_rat = mean_abs_log2_aFC),
               by = "gene_id") |>
    inner_join(gene_afc |> rename(mean_abs_log2_aFC_gtex = mean_abs_log2_aFC),
               by = c("human_id" = "gene_id"))
```

Spearman correlation:

```{r}
with(d, cor(mean_abs_log2_aFC_gtex, mean_abs_log2_aFC_rat, method = "spearman"))
d |>
    ggplot(aes(x = mean_abs_log2_aFC_gtex, y = mean_abs_log2_aFC_rat)) +
    geom_point(size = 0.25) +
    stat_smooth()
```

As seen in boxplots above, magnitude of aFC for top eQTLs is lower in HS rat than in GTEx. Is this because more eQTLs were found in GTEx? Subsample to the most significant N eQTLs for all tissues (where N is smallest count among tissues) and compare:

```{r}
N_min <- both |>
    count(tissue, sort = TRUE) |>
    slice(n()) |>
    pull(n)
both |>
    group_by(tissue) |>
    arrange(qval) |>
    slice(1:N_min) |>
    ungroup() |>
    ggplot(aes(x = tissue, y = log2_aFC)) +
    geom_boxplot(outlier.size = 0.25) +
    coord_flip()
```

## Compare against correlation with other tissues (cultured cells)

```{r}
gtex_cell <- tibble(file = list.files("../../data/gtex/GTEx_Analysis_v8_eQTL",
                                 pattern = "Cells_.+",
                                 full.names = TRUE)) |>
    mutate(tissue = str_match(file, "Cells_(.+)\\.v8")[, 2]) |>
    reframe(
        read_tsv(file, col_types = cols(gene_id = "c", qval = "d", pval_beta = "-",
                                        log2_aFC = "d", .default = "-")),
        .by = c(file, tissue)
    ) |>
    filter(qval <= 0.05) |>  # GTEx site says to do <=
    select(-file)
both_cell <- bind_rows(
    eqtls |> mutate(tissue = str_c("HSrat_", tissue)),
    gtex_cell |> mutate(tissue = str_c("GTEx_", tissue))
)

gene_afc_cell <- both_cell |>
    group_by(gene_id) |>
    summarise(mean_abs_log2_aFC = mean(abs(log2_aFC)))

d <- gene_map |>
    inner_join(gene_afc_cell |> rename(mean_abs_log2_aFC_rat = mean_abs_log2_aFC),
               by = "gene_id") |>
    inner_join(gene_afc_cell |> rename(mean_abs_log2_aFC_gtex = mean_abs_log2_aFC),
               by = c("human_id" = "gene_id"))
```

Spearman correlation:

```{r}
with(d, cor(mean_abs_log2_aFC_gtex, mean_abs_log2_aFC_rat, method = "spearman"))
d |>
    ggplot(aes(x = mean_abs_log2_aFC_gtex, y = mean_abs_log2_aFC_rat)) +
    geom_point(size = 0.25) +
    stat_smooth()
```

## Compare against correlation with all non-brain tissues

```{r}
gtex_other <- tibble(file = list.files("../../data/gtex/GTEx_Analysis_v8_eQTL",
                                     full.names = TRUE)) |>
    filter(str_detect(file, "Brain", negate = TRUE)) |>
    mutate(tissue = str_match(file, "eQTL/(.+)\\.v8")[, 2]) |>
    reframe(
        read_tsv(file, col_types = cols(gene_id = "c", qval = "d", pval_beta = "-",
                                        log2_aFC = "d", .default = "-")),
        .by = c(file, tissue)
    ) |>
    filter(qval <= 0.05) |>  # GTEx site says to do <=
    select(-file)
both_other <- bind_rows(
    eqtls |> mutate(tissue = str_c("HSrat_", tissue)),
    gtex_other |> mutate(tissue = str_c("GTEx_", tissue))
)

gene_afc_other <- both_other |>
    group_by(gene_id) |>
    summarise(mean_abs_log2_aFC = mean(abs(log2_aFC)))

d <- gene_map |>
    inner_join(gene_afc_other |> rename(mean_abs_log2_aFC_rat = mean_abs_log2_aFC),
               by = "gene_id") |>
    inner_join(gene_afc_other |> rename(mean_abs_log2_aFC_gtex = mean_abs_log2_aFC),
               by = c("human_id" = "gene_id"))
```

Spearman correlation:

```{r}
with(d, cor(mean_abs_log2_aFC_gtex, mean_abs_log2_aFC_rat, method = "spearman"))
d |>
    ggplot(aes(x = mean_abs_log2_aFC_gtex, y = mean_abs_log2_aFC_rat)) +
    geom_point(size = 0.25) +
    stat_smooth()
```
